Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-05-03 manuscript.tex
latest

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, density is actually high relative to carrying capacity of the stressful environment \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}. Thus, even though $b\ll 1$, we can still have $b/d\gg 1$ and $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.

Second, some mutations are more common than others, which can change the direction of evolution \citep{yampolsky_2001}. In particular, very few mutations can perform difficult trait changes such as increasing resilience to intense physical disturbance. Quantifying which trait changes are more mutationally accessible than others is beyond the scope of this article, and so we will assume that $b$, $c$ and $d$ all have the same distribution of beneficial fitness effects.

Finally, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$). Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.

>>>>

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, density is actually high relative to carrying capacity of the stressful environment \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}. Thus, even though $b\ll 1$, we still have $b/d\gg 1$ and $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others, which can change the direction of evolution \citep{yampolsky_2001}. Quantifying how traits are affected by mutations is beyond the scope of this article, and so we will not evaluate either of these effects: mutations affect $b$, $c$ and $d$ independently, and they all have the same distribution of beneficial fitness effects.

The third effect is that selection can favor some trait changes over others. The probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$). Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.
